---
title: "Plot EC-annotated tumor-enriched splice junctions"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Ryan Corbett
date: 2026
params:
  plot_ci: TRUE
---

Load required packages

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(data.table)
  library(rprojroot)
  library(R.utils)
  library(ggsci)
  library(ggforce)
  library(ggpubr)
})
```

Set directory paths

```{r}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "04-summarize-TEJs")
input_dir <- file.path(analysis_dir, "input")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}

source(file.path(root_dir, "figures", "theme_for_plots.R"))
```

Set file paths

```{r}
tej_file <- file.path(root_dir, "analyses",
                      "04-summarize-TEJs",
                      "results",
                      "recurrent-primary-tumor-enriched-oncofetal-splice-junctions-annotated.tsv.gz")

uniprot_file <- file.path(root_dir, "analyses",
                        "pepMLM-preprocessing",
                        "input",
                        "uniprotkb_organism_id_9606_AND_reviewed_2025_06_18.txt")
```

Wrangle data

```{r}
tej_df <- read_tsv(tej_file)

uniprot_df <- read_tsv(uniprot_file)
```

Current uniprot topological domain annotations include associations with ANY membrane (ER, mitochondria), not just cell membrane.
Therefore, we need to identify proteins that localize to cell membranes and update annotations. 
These will include keywords "Membrane"/"cell membrane", and "single-pass"/"multi-pass"

```{r}
# extract known extracellular proteins based on keywords
cell_membrane_proteins <- uniprot_df %>%
  dplyr::filter(
    grepl("Single-pass type I membrane protein|Multi-pass membrane protein|Single-pass membrane protein|Single-pass type II membrane protein|Single-pass type III membrane protein|Single-pass type IV membrane protein", `Subcellular location [CC]`),
    grepl("cell membrane|Cell membrane", `Subcellular location [CC]`),
    # remove ER membrane 
    !grepl("Endoplasmic reticulum membrane; Single-pass type I membrane protein|endoplasmic reticulum (ER) membrane", `Subcellular location [CC]`),
    !is.na(`Gene Names (primary)`)
                ) %>%
  pull(`Gene Names (primary)`)
```

Update TEJ df uniprot annotation to remove annotations for proteins that don't localize to extracellular membrane

```{r}
tej_df <- tej_df %>% 
  dplyr::mutate(uniprot_domain_type = case_when(
    gene_symbol %in% cell_membrane_proteins & !is.na(uniprot_domain_type) ~ uniprot_domain_type,
    TRUE ~ NA_character_
  )) %>%
  dplyr::mutate(uniprot_domain_start = case_when(
    is.na(uniprot_domain_type) ~ NA,
    TRUE ~ uniprot_domain_start
  ),
  uniprot_domain_end = case_when(
    is.na(uniprot_domain_type) ~ NA,
    TRUE ~ uniprot_domain_end
    )) %>%
  dplyr::mutate(consequence = case_when(
    grepl("EC", uniprot_domain_type) & gene_symbol %in% cell_membrane_proteins ~ "EC domain",
    !is.na(pfam_name) ~ "Other functional domain",
    in_cds == "Yes" ~ "Other coding sequence",
    TRUE ~ "Putative non-coding"
  ))
```

Determine distribution of junctions affecting extracellular and pfam domains

```{r}
consequence_df <- tej_df %>%
  distinct(junction, gene_symbol,
           consequence, pfam_name, uniprot_domain_type, in_cds) %>%
  dplyr::mutate(consequence = case_when(
    grepl("EC", uniprot_domain_type) & gene_symbol %in% cell_membrane_proteins ~ "EC domain",
    !is.na(pfam_name) ~ "Other functional domain",
    in_cds == "Yes" ~ "Other coding sequence",
    TRUE ~ "Putative non-coding"
  ))

table(consequence_df$consequence)

consequence_ct <- consequence_df %>%
  count(consequence) %>%
  arrange(n) %>%
  dplyr::mutate(consequence = fct_relevel(consequence,
                                          unique(consequence)))

ggplot(consequence_ct,
       aes(y = consequence, x = n, fill = consequence)) +
  geom_bar(stat = "identity", color = "black", show.legend = FALSE) + 
  theme_Publication() +
  labs(y = NULL,
     x = "N TEJs") +
  scale_fill_npg()

ggsave(file.path(plot_dir,
                 "tej-n-by-consequence.pdf"),
       width = 4.5, height = 2.5)
```

Plot EC-annotated TEJs by histology

```{r}
groups <- sort(unique(tej_df$plot_group))

file_names <- c("ATRT", "CPT", "CRANIO",
                "DMG", "EPN", "GCT",
                "LGG", "MB", "MNG",
                "MES", "GNT", "NFP",
                "NNT", "OLIGO", "CNSemb", 
                "HGG", "Other", "SWN")
names(file_names) <- groups

preference_palette <- c("EI" = "#56B4E9",
                        "ES" = "#0072B2",
                        "RI" = "#009E73",
                        "A3-" = "#E69F00",
                        "A3+" = "#F0E442",
                        "A5-" = "#D55E00",
                        "A5+" = "#CC79A7")

for (hist in groups){

  recurrent_ec_events <- tej_df %>%
    dplyr::filter(plot_group == hist,
                  grepl("EC domain", consequence) & gene_symbol %in% cell_membrane_proteins) %>%
    dplyr::count(gene_symbol, preference_code) %>%
    dplyr::filter(n >= 5) %>%
    arrange(desc(gene_symbol)) %>%
    dplyr::mutate(gene_symbol = fct_relevel(gene_symbol,
                                            unique(gene_symbol)))
  
  if (nrow(recurrent_ec_events) > 0){

    recurrent_ec_events %>%
      
    ggplot(aes(y = gene_symbol, x = n, fill = preference_code)) +
        geom_bar(stat="identity", position="stack", color = "black", linewidth = 0.2) + 
        labs(x = "Count",
             y = NULL,
             fill = "Event type",
             title = hist) + 
        scale_fill_manual(values = preference_palette, 
             breaks = sort(names(preference_palette))) +
        theme_Publication()
  
    ggsave(file.path(plot_dir, glue::glue("{file_names[hist]}-recurrent-ec-tejs-by-gene.pdf")),
           height = 12, width = 5)
    
  }
  
}
```

Plot high recurrence ATRT, HGG EC-localized events only

```{r}
hgg_atrt_ec_events <- tej_df %>%
  dplyr::filter(plot_group %in% c("Other high-grade glioma",
                                  "DIPG or DMG",
                                  "Atypical Teratoid Rhabdoid Tumor"),
                  grepl("EC domain", consequence) & gene_symbol %in% cell_membrane_proteins) %>%
  dplyr::count(gene_symbol, preference_code) %>%
  dplyr::filter(n >= 20) %>%
  pull(gene_symbol)

hgg_atrt_ec_events_df <- tej_df %>%
  dplyr::mutate(plot_group = case_when(
    plot_group %in% c("DIPG or DMG",
                      "Other high-grade glioma") ~ "High-grade glioma",
    TRUE ~ plot_group
  )) %>%
  dplyr::filter(grepl("EC domain", consequence) & gene_symbol %in% cell_membrane_proteins,
                gene_symbol %in% hgg_atrt_ec_events,
                  plot_group %in% c("High-grade glioma",
                                    "Atypical Teratoid Rhabdoid Tumor")) %>%
  dplyr::count(gene_symbol, preference_code, plot_group) %>%
  arrange(desc(gene_symbol)) %>%
  dplyr::mutate(gene_symbol = fct_relevel(gene_symbol,
                                          unique(gene_symbol)))
  
hgg_atrt_ec_events_df %>%
  
  ggplot(aes(y = gene_symbol, x = n, fill = preference_code)) +
      geom_bar(stat="identity", position="stack", color = "black", linewidth = 0.2) + 
      labs(x = "Count",
           y = NULL,
           fill = "Event type") + 
    facet_wrap(~plot_group, nrow = 1, scale = "free",
               labeller = labeller(plot_group = label_wrap_gen(16))) +
      scale_fill_manual(values = preference_palette, 
           breaks = sort(names(preference_palette))) +
      theme_Publication()

  ggsave(file.path(plot_dir, "atrt-hgg-recurrent-ec-tejs-by-gene.pdf"),
         height = 9, width = 8)
```

Save TEJ data frame with updated domain annotation

```{r}
write_tsv(tej_df,
          file.path(results_dir,
                    "recurrent-primary-tumor-enriched-oncofetal-splice-junctions-annotated-domain-updated.tsv.gz"))
```

print session info
```{r}
sessionInfo()
```