---
title: "Generate TEJ-associated splice event PSI matrix"
author: "Ryan Corbett"
date: "2026"
output:
  html_document:
    toc: true
    df_print: paged
  html_notebook:
    toc: true
    toc_float: true
params:
  plot_ci: true
---

### Script objectives
* Obtains differential splice event associated with each TEJ 
* Loads rMATS files to obtain PSIs (IncLevel1 values) for each TEJ-associated splice event
* Merges PSIs and generates PSI matrix  

load necessary packages
```{r}
library(tidyverse)
library(data.table)
library(gtools)
library(ggpubr)
library(ggsci)
```

Set dir paths
```{r}
# create directories
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "04-summarize-TEJs")
input_dir <- file.path(analysis_dir, "input")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

source(file.path(root_dir, "figures",
                 "theme_for_plots.R"))
```

Set file paths
```{r}
tej_file <- file.path(root_dir, "analyses",
                      "04-summarize-TEJs",
                      "results",
                      "recurrent-primary-tumor-enriched-oncofetal-splice-junctions-annotated-domain-updated.tsv.gz")


splice_event_file <- file.path(root_dir, "analyses",
                               "03-tesj-differential-splicing",
                               "results",
                               "tumor-enriched-oncofetal-junction-diff-splice-event-annotation.tsv.gz")

cohort_hist_file <- file.path(root_dir, 
                              "analyses",
                              "00-create-cohort-histologies",
                              "results",
                              "cohort-histologies.tsv")
```

Wrangle data
```{r}
tej_df <- read_tsv(tej_file)

cohort_hist <- read_tsv(cohort_hist_file)

tej_splice_event_key <- read_tsv(splice_event_file) %>%
  dplyr::filter(junction %in% tej_df$junction,
                event_type_sample == consensus_jc_event_type) %>%
  dplyr::count(junction, splice_id, consensus_jc_event_type) %>%
  dplyr::arrange(desc(n)) %>%
  distinct(junction, .keep_all = TRUE)
```

Obtain splice event PSIs from each PBTA rMATS file
```{r}
a5ss_splice_event_df <- qs2::qs_read(file.path(data_dir,
                                               "pbta-rmats_merged_raw_A5SS.qs2")) %>% 
  dplyr::mutate(sample_id = str_remove(sample_id, "_[^_]+$"),
                splice_id = glue::glue("{chr}:{longExonStart_0base + 1}-{longExonEnd}_{shortES + 1}-{shortEE}_{flankingES + 1}-{flankingEE}_{strand}")) %>%
  dplyr::filter(splice_id %in% tej_splice_event_key$splice_id,
                sample_id %in% cohort_hist$Kids_First_Biospecimen_ID) %>%
  dplyr::select(sample_id, splice_id, IncLevel1)

a3ss_splice_event_df <- qs2::qs_read(file.path(data_dir,
                                               "pbta-rmats_merged_raw_A3SS.qs2")) %>% 
  dplyr::mutate(sample_id = str_remove(sample_id, "_[^_]+$"),
                splice_id = glue::glue("{chr}:{longExonStart_0base + 1}-{longExonEnd}_{shortES + 1}-{shortEE}_{flankingES + 1}-{flankingEE}_{strand}")) %>%
  dplyr::filter(splice_id %in% tej_splice_event_key$splice_id,
                sample_id %in% cohort_hist$Kids_First_Biospecimen_ID) %>%
  dplyr::select(sample_id, splice_id, IncLevel1)

ri_splice_event_df <- qs2::qs_read(file.path(data_dir,
                                               "pbta-rmats_merged_raw_RI.qs2")) %>% 
  dplyr::mutate(sample_id = str_remove(sample_id, "_[^_]+$"),
                splice_id = glue::glue("{chr}:{riExonStart_0base}-{riExonEnd}_{upstreamES + 1}-{upstreamEE}_{downstreamES + 1}-{downstreamEE}_{strand}")) %>%
  dplyr::filter(splice_id %in% tej_splice_event_key$splice_id,
                sample_id %in% cohort_hist$Kids_First_Biospecimen_ID) %>%
  dplyr::select(sample_id, splice_id, IncLevel1)

se_splice_event_df <- qs2::qs_read(file.path(data_dir,
                                               "pbta-rmats_merged_raw_SE.qs2")) %>% 
  dplyr::mutate(sample_id = str_remove(sample_id, "_[^_]+$"),
                splice_id = glue::glue("{chr}:{exonStart_0base + 1}-{exonEnd}_{upstreamES + 1}-{upstreamEE}_{downstreamES + 1}-{downstreamEE}_{strand}")) %>%
  dplyr::filter(splice_id %in% tej_splice_event_key$splice_id,
                sample_id %in% cohort_hist$Kids_First_Biospecimen_ID) %>%
  dplyr::select(sample_id, splice_id, IncLevel1)
```

Create splice event PSI matrix
```{r}
merged_splice_event_mat <- se_splice_event_df %>%
  bind_rows(a5ss_splice_event_df,
            a3ss_splice_event_df,
            ri_splice_event_df) %>%
  pivot_wider(id_cols = splice_id,
               names_from = "sample_id",
               values_from = "IncLevel1")
```

Write TEJ-splice event key and PSI matrix to output
```{r}
write_tsv(tej_splice_event_key,
          file.path(results_dir,
                    "recurrent-primary-tej-associated-splice-events.tsv"))

saveRDS(merged_splice_event_mat,
          file.path(results_dir,
                    "recurrent-primary-tumor-enriched-oncofetal-splice-event-psis.rds"))
```

Print session info
```{r}
sessionInfo()
```