---
title: "Summarize tumor-enriched splicing"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Ryan Corbett
date: 2025
params:
  plot_ci: TRUE
---

This scripts identifies recurrent tumor-enriched junctions in primary PBTA tumors and generates summary plots by sample and tumor histology

Load required packages

```{r}
suppressPackageStartupMessages({
  library(qs2)
  library(tidyverse)
  library(data.table)
  library(rprojroot)
  library(R.utils)
  library(ggsci)
  library(ggforce)
  library(ggpubr)
})
```

Set directory paths

```{r}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "05-summarize-TEJs")
input_dir <- file.path(analysis_dir, "input")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}

source(file.path(root_dir, "figures", "theme_for_plots.R"))
```

Set file paths

```{r}
tej_file <- file.path(root_dir, "analyses",
                      "04-tej-differential-splicing",
                      "results",
                      "tumor-enriched-oncofetal-diff-splice-junctions.qs2")

hist_file <- file.path(root_dir, "analyses",
                       "00-create-cohort-histologies",
                       "results",
                       "cohort-histologies.tsv")
```

Wrangle data

```{r}
tej_df <- qs2::qs_read(tej_file)

hist <- read_tsv(hist_file)
```

Filter TEJ df for samples in cohort hist, and append relevant metadata

```{r}
tej_df <- tej_df %>%
  dplyr::filter(Kids_First_Biospecimen_ID %in% hist$Kids_First_Biospecimen_ID) %>%
  left_join(hist %>% 
              dplyr::select(Kids_First_Biospecimen_ID,
                            sample_type,
                            tumor_descriptor,
                            composition,
                            molecular_subtype,
                            RNA_library,
                            plot_group,
                            plot_group_hex),
            by = "Kids_First_Biospecimen_ID")
```

Filter TEJs for those in primary independent specimens

```{r}
primary_independent_specimens <- hist %>%
  dplyr::filter(is_independent_primary == "Yes") %>%
  pull(Kids_First_Biospecimen_ID)

primary_tej_df <- tej_df %>%
  dplyr::filter(Kids_First_Biospecimen_ID %in% primary_independent_specimens)
```

Identify most recurrent exon-exon junction for each TEJ boundary (prioritizing those that are most recurrent in primary tumors), and filter full and primary df for these TEJs

```{r}
# get most recurrent exon-exon jc for boundaries in primary tumors
top_primary_bounds <- primary_tej_df %>%
  # count TEJ N per boundary
  dplyr::count(boundary, junction) %>%
  # arrange in descending n order
  arrange(desc(n)) %>%
  # keep most prevalent TEJ per boundary
  distinct(boundary, .keep_all = TRUE) %>%
  pull(junction)

# for TEJs in non-primary tumors only, get most recurrent exon-exon jc
top_other_bounds <- tej_df %>%
  dplyr::filter(!boundary %in% primary_tej_df$boundary) %>%
  # count TEJ N per boundary
  dplyr::count(boundary, junction) %>%
  # arrange in descending n order
  arrange(desc(n)) %>%
  # keep most prevalent TEJ per boundary
  distinct(boundary, .keep_all = TRUE) %>%
  pull(junction)

# filter TEJ df
tej_df <- tej_df %>%
  dplyr::filter(junction %in% c(top_primary_bounds,
                                top_other_bounds)) 

# save to output
write_tsv(tej_df,
          file.path(results_dir,
                    "tumor-enriched-oncofetal-splice-junctions-cohort-filtered.tsv.gz"))

primary_tej_df <- primary_tej_df %>%
  dplyr::filter(junction %in% c(top_primary_bounds,
                                top_other_bounds)) 
```


Filter for recurrent events (i.e., event observed in >5% of tumor histologies)

```{r}
# Get number of patients per histology
# n_per_hist <- hist %>%
#   dplyr::filter(is_independent_primary == "Yes") %>%
#   distinct(Kids_First_Biospecimen_ID, .keep_all = TRUE) %>%
#   count(plot_group) %>%
#   dplyr::rename(n_hist = n)

primary_tej_df <- primary_tej_df %>%
  dplyr::mutate(plot_group_collapsed = case_when(
    plot_group %in% c("Other high-grade glioma",
                      "DIPG or DMG") ~ "HGG",
      TRUE ~ plot_group
  ))

# obtain frequency of splice events by hist
n_per_hist_jc <- primary_tej_df %>%
  count(plot_group_collapsed, junction) %>%
  dplyr::rename(n_hist_jc = n) %>%
  # filter for events in >=3 samples in histology
  dplyr::filter(n_hist_jc >= 3)

# retain only recurrent events
recurrent_primary_tej_df <- primary_tej_df %>%
  dplyr::filter(glue::glue("{plot_group_collapsed}-{junction}") %in% glue::glue("{n_per_hist_jc$plot_group_collapsed}-{n_per_hist_jc$junction}")) %>%
  dplyr::select(-plot_group_collapsed)
```

Plot total and recurrent tej n by histology

```{r}
total_tej_hist_n <- primary_tej_df %>%
  distinct(plot_group, junction) %>%
  dplyr::count(plot_group) %>%
  dplyr::rename(total_tej_n = n)

recurrent_tej_hist_n <- recurrent_primary_tej_df %>%
  distinct(plot_group, junction) %>%
  dplyr::count(plot_group) %>%
  dplyr::rename(recurrent_tej_n = n) %>%
  dplyr::arrange(recurrent_tej_n)

# order based on N recurrent events
hist_order <- recurrent_tej_hist_n$plot_group

tej_n_plot_df <- total_tej_hist_n %>%
  left_join(recurrent_tej_hist_n) %>%
  dplyr::mutate(nonrecurrent_tej_n = total_tej_n - recurrent_tej_n) %>%
  dplyr::select(-total_tej_n) %>%
  pivot_longer(-plot_group,
               names_to = "tej_group",
               values_to = "n") %>%
  dplyr::mutate(tej_group = case_when(
    tej_group == "recurrent_tej_n" ~ "Recurrent",
    tej_group == "nonrecurrent_tej_n" ~ "Non-recurrent"
  )) %>%
  dplyr::mutate(plot_group = fct_relevel(plot_group,
                                         hist_order)) %>%
  dplyr::arrange(plot_group)

# generate plot
tej_n_barplot <- tej_n_plot_df %>%
  ggplot(aes(y = factor(plot_group), x = n, fill = tej_group)) +
  geom_bar(stat = "identity", color = "black", show.legend = TRUE) + 
  scale_fill_npg() +
  theme_Publication() +
  labs(x = "N TEJs",
     y = NULL,
     fill = NULL)

# save to output
ggsave(file.path(plot_dir,
                 "tej-n-by-hist-barplot.pdf"),
       tej_n_barplot,
       width = 7, height = 6)

tej_n_barplot
```

Plot number of total TEJs by sample and histology

```{r}
# obtain number of events by sample
hist_tej_ct <- primary_tej_df %>%
  count(plot_group, Kids_First_Biospecimen_ID) %>%
  dplyr::rename(tej_n = n) %>%
  dplyr::mutate(plot_group = fct_relevel(plot_group,
                                        unique(plot_group)))

# define plot group palette
plotgroup_palette <- unique(hist$plot_group_hex)
names(plotgroup_palette) <- unique(hist$plot_group)
plotgroup_palette <- plotgroup_palette[!is.na(names(plotgroup_palette))]

plotgroup_palette["Non-neoplastic tumor"] <- "black"

total_tej_n_plot <- hist_tej_ct %>%
  dplyr::mutate(plot_group = fct_relevel(plot_group,
                                        as.character(hist_order))) %>%
  
  ggplot(aes(y = plot_group, x = tej_n)) +
  geom_sina(aes(color = plot_group), shape = 20, 
              width = 0.1, size = 2, alpha = 0.5,
              show.legend = FALSE) +
    geom_boxplot(fill = NA, alpha = 0.3, size = 0.4, outlier.shape = NA,
               show.legend = FALSE) +
  scale_color_manual(values = plotgroup_palette, 
                     breaks = sort(names(plotgroup_palette))) +
  theme_Publication() +
  labs(x = "N TEJs/sample",
     y = NULL)

ggsave(file.path(plot_dir,
                 "total-tej-n-by-sample-hist.pdf"),
       total_tej_n_plot,
       width = 5, height = 6)

total_tej_n_plot
```

Plot number of recurrent TEJs by sample and histology

```{r}
set.seed(2025)
# obtain number of events by sample
histology_jc_ct <- recurrent_primary_tej_df %>%
  count(plot_group, Kids_First_Biospecimen_ID) %>%
  dplyr::rename(jc_n = n) %>%
  dplyr::mutate(plot_group = fct_relevel(plot_group,
                                        unique(plot_group)))

recurrent_tej_n_plot <- histology_jc_ct %>%
  dplyr::mutate(plot_group = fct_relevel(plot_group,
                                        as.character(hist_order))) %>%
  
  ggplot(aes(y = plot_group, x = jc_n)) +
  geom_sina(aes(color = plot_group), shape = 20, 
              width = 0.1, size = 2, alpha = 0.5,
              show.legend = FALSE) +
    geom_boxplot(fill = NA, alpha = 0.3, size = 0.4, outlier.shape = NA,
               show.legend = FALSE) +
  scale_color_manual(values = plotgroup_palette, 
                     breaks = sort(names(plotgroup_palette))) +
  xlim(c(0, 1300)) +
  theme_Publication() +
  labs(x = "N recurrent TEJs",
     y = NULL)

ggsave(file.path(plot_dir,
                 "recurrent-tej-n-by-sample-hist.pdf"),
       recurrent_tej_n_plot,
       width = 4.5, height = 6)

recurrent_tej_n_plot
```


Plot distribution of TEJ splice event types by histology

```{r}
recurrent_primary_tej_df <- recurrent_primary_tej_df %>%
  dplyr::mutate(preference_code = case_when(
    consensus_jc_event_type == "exon skipping" ~ "ES",
    consensus_jc_event_type == "exon inclusion" ~ "EI",
    consensus_jc_event_type == "intron retention" ~ "RI",
    consensus_jc_event_type == "A3SS+" ~ "A3+",
    consensus_jc_event_type == "A3SS-" ~ "A3-",
    consensus_jc_event_type == "A5SS+" ~ "A5+",
    consensus_jc_event_type == "A5SS-" ~ "A5-"
  ))

# define tumor preference palette
preference_palette <- c("EI" = "#56B4E9",
                        "ES" = "#0072B2",
                        "RI" = "#009E73",
                        "A3-" = "#E69F00",
                        "A3+" = "#F0E442",
                        "A5-" = "#D55E00",
                        "A5+" = "#CC79A7")

unique_hist_event_ct <- recurrent_primary_tej_df %>%
  distinct(plot_group, junction) %>%
  dplyr::count(plot_group) %>%
  dplyr::rename(hist_event_n = n)

hist_event_type_ct <- recurrent_primary_tej_df %>%
  distinct(junction, plot_group, .keep_all = TRUE) %>%
  dplyr::count(plot_group, preference_code) %>%
  left_join(unique_hist_event_ct) %>%
  dplyr::mutate(n_norm = n/hist_event_n)

event_type_freq_barplot <- hist_event_type_ct %>%
  dplyr::mutate(plot_group = fct_relevel(plot_group,
                                        as.character(hist_order))) %>%
  ggplot(aes(y = plot_group, x = n_norm, fill = preference_code)) +
  geom_bar(stat = "identity", color = "black", show.legend = TRUE) + 
  scale_fill_manual(values = preference_palette, 
                     breaks = sort(names(preference_palette))) +
  theme_Publication() +
  labs(x = "Proportion TEJs",
     y = NULL,
     fill = NULL)

ggsave(file.path(plot_dir,
                 "tejs-by-event-type-hist-barplot.pdf"),
       event_type_freq_barplot,
       width = 6, height = 6)

event_type_freq_barplot
```

Plot frequency of tumor-enriched and oncofetal TEJs by histology

```{r}
hist_tej_spec_ct <- recurrent_primary_tej_df %>%
  distinct(junction, plot_group, junction_preference) %>%
  count(plot_group, junction_preference) %>%
  left_join(unique_hist_event_ct) %>%
  dplyr::mutate(n_norm = n/hist_event_n)

tej_spec_barplot <- hist_tej_spec_ct %>%
  dplyr::mutate(plot_group = fct_relevel(plot_group,
                                        as.character(hist_order))) %>%
  ggplot(aes(y = plot_group, x = n_norm, fill = junction_preference)) +
  geom_bar(stat = "identity", color = "black", show.legend = TRUE) + 
  scale_fill_npg() +
  theme_Publication() +
  labs(x = "Proportion TEJs",
     y = NULL,
     fill = NULL)

ggsave(file.path(plot_dir,
                 "tej-by-specificity-hist-barplot.pdf"),
       tej_spec_barplot,
       width = 7, height = 6)

tej_spec_barplot
```

Save recurrent splice event df to output

```{r}
write_tsv(recurrent_primary_tej_df,
          file.path(results_dir,
          "recurrent-primary-tumor-enriched-oncofetal-splice-junctions.tsv.gz"))
```

Print session info

```{r}
sessionInfo()
```
