---
title: "TEJ transcript annotation"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Ryan Corbett
date: 2025
params:
  plot_ci: TRUE
---

Load required packages

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(data.table)
  library(rprojroot)
  library(R.utils)
  library(ggsci)
  library(ggforce)
  library(ggpubr)
})
```

Set directory paths

```{r}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "05-summarize-TEJs")
input_dir <- file.path(analysis_dir, "input")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}

source(file.path(root_dir, "figures", "theme_for_plots.R"))
source(file.path(analysis_dir, "util", "annotation_functions.R"))
```

Set file paths

```{r}
tej_file <- file.path(results_dir,
                       "recurrent-primary-tumor-enriched-oncofetal-splice-junctions.tsv.gz")

gtf_file <- file.path(data_dir,
                      "gencode.v39.primary_assembly.annotation.gtf.gz")
```

Wrangle data

```{r}
tej_df <- read_tsv(tej_file)

gtf <- rtracklayer::import(gtf_file)

gtf_df <- as.data.frame(gtf) %>%
  dplyr::filter(type == "exon") 
```

Determine if TEJ exons map to GTF-annotated exons

```{r}
starts <- gtf_df %>%
  distinct(seqnames, start, end, .keep_all = TRUE) %>%
  dplyr::mutate(chr_start = glue::glue("{seqnames}:{start}")) %>%
  pull(chr_start)

ends <- gtf_df %>%
  distinct(seqnames, start, end, .keep_all = TRUE) %>%
  dplyr::mutate(chr_start = glue::glue("{seqnames}:{end}")) %>%
  pull(chr_start)

tej_annot_df <- tej_df %>%
  # retain unique TEJs
  distinct(junction, gene_symbol, preference_code, chr,
           up_jc_start, up_jc_end, down_jc_start, down_jc_end,
           strand) %>%
  dplyr::mutate(chr = case_when(
    grepl("KI", chr) ~ str_remove(chr, "chr"),
    TRUE ~ chr
  )) %>%
  # define exon coords for GTF mapping
  dplyr::mutate(up_bound = glue::glue("{chr}:{up_jc_end}"),
                down_bound = glue::glue("{chr}:{down_jc_start}")) %>%
  # determine if junction bounds are annotated
  dplyr::mutate(up_bound_annotated = case_when(
    up_bound %in% ends ~ "Yes",
    TRUE ~ "No"
  )) %>%
  dplyr::mutate(down_bound_annotated = case_when(
    down_bound %in% starts ~ "Yes",
    TRUE ~ "No"
  ))

# if exon-exon boundary and one boundary not annotated, classify as novel
tej_annot_df <- tej_annot_df %>%
  dplyr::mutate(novel_ss = case_when(
    preference_code == "RI" & up_bound_annotated == "No" & down_bound_annotated == "No" ~ "Yes",
    preference_code == "RI" ~ "No",
    up_bound_annotated == "No" | down_bound_annotated == "No" ~ "Yes",
    TRUE ~ "No"
  ))
```

Run `annotate_junctions()` function to annotate junctions to transcripts, exons, and introns

```{r}
# Subset gtf to make for easier filtering within function
gtf_sub_dt <- gtf_df %>%
  dplyr::filter(start %in% tej_annot_df$down_jc_start | end %in% tej_annot_df$up_jc_end) %>%
  data.table::as.data.table()

# Create the splice-boundary lookup indices once. These are reused for every
# junction annotation instead of sorting/copying the GTF within each function call.
data.table::setindexv(gtf_sub_dt, c("seqnames", "gene_name", "end"))
data.table::setindexv(gtf_sub_dt, c("seqnames", "gene_name", "start"))

# run annotation
system.time({
  tej_annot_df <- tej_annot_df %>%
    dplyr::mutate(
      junction_name = suppressMessages(pmap_chr(
          pick(chr,
               up_jc_start,
               up_jc_end,
               down_jc_start,
               down_jc_end,
               gene_symbol,
               preference_code),
          \(...) annotate_junctions(
            chr = ..1,
            up_exon_start = ..2,
            up_exon_end = ..3,
            down_exon_start = ..4,
            down_exon_end = ..5,
            gene_symbol = ..6,
            preference = ..7,
            gtf = gtf_sub_dt
          )
        )
      )
    )
})
```

Join annotation to recurrent TEJ data frame

```{r}
tej_df <- tej_df %>%
  left_join(tej_annot_df %>%
              distinct(junction,
                      novel_ss,
                      junction_name)) %>%
  dplyr::mutate(
    up_exon = case_when(
      grepl(":exon", junction_name) ~ sapply(
        strsplit(junction_name, ", |:"),
        function(x) if (length(x) >= 2) x[2] else NA_character_
      ),
      TRUE ~ NA_character_
    )
  ) %>%
  dplyr::mutate(
    down_exon = case_when(
      grepl(":exon", junction_name) ~ sapply(
        strsplit(junction_name, ", |:"),
        function(x) if (length(x) >= 2) x[3] else NA_character_
      ),
      TRUE ~ NA_character_
    )
  ) %>%
  dplyr::mutate(up_exon = as.numeric(str_remove(up_exon, "exon")),
                down_exon = as.numeric(str_remove(down_exon, "exon"))) %>%
  dplyr::mutate(status = case_when(
    grepl("novel", junction_name) ~ "Novel splice site",
    !grepl(":exon|, intron|:intron", junction_name) ~ "Novel junction",
    !is.na(up_exon) & !is.na(down_exon) & up_exon != down_exon + 1 & up_exon + 1 != down_exon ~ "Novel junction",
    TRUE ~ "Annotated junction"
  ))
```

Plot novel splice site usage frequency among tejs by histology

```{r}
unique_hist_tej_ct <- tej_df %>%
  distinct(plot_group, junction) %>%
  dplyr::count(plot_group) %>%
  dplyr::rename(hist_tej_n = n)

hist_tej_novelss_ct <- tej_df %>%
  distinct(junction, plot_group, .keep_all = TRUE) %>%
  dplyr::count(plot_group, status) %>%
  left_join(unique_hist_tej_ct) %>%
  dplyr::mutate(n_norm = n/hist_tej_n)

hist_order <- c("Other CNS embryonal tumor", 
                "Pineoblastoma", "Oligodendroglioma",
                "Neurofibroma plexiform", "Mesenchymal tumor", 
                "Rare CNS tumor", "Germ cell tumor",
                "Meningioma", "Schwannoma/MPNST", 
                "Non-neoplastic tumor", "Craniopharyngioma", 
                "Mixed neuronal-glial tumor",
                "Atypical Teratoid Rhabdoid Tumor", 
                "Choroid plexus tumor", "Ependymoma",
                "DIPG or DMG", "Other high-grade glioma",
                "Low-grade glioma", "Medulloblastoma")

# generate plot
novelss_freq_barplot <- hist_tej_novelss_ct %>%
  dplyr::filter(status != "Annotated junction") %>%
  dplyr::mutate(plot_group = fct_relevel(plot_group,
                                        as.character(hist_order))) %>%
  ggplot(aes(y = plot_group, x = n_norm, fill = status)) +
  geom_bar(stat = "identity", color = "black", show.legend = TRUE) + 
  scale_fill_npg() +
  theme_Publication() +
  labs(x = "Proportion TEJs",
     y = NULL,
     fill = "Junction annotation")

ggsave(file.path(plot_dir,
                 "tejs-by-junction-status-hist-barplot.pdf"),
       novelss_freq_barplot,
       width = 8, height = 6)
```

Write annotation results to output

```{r}
write_tsv(tej_df,
          file.path(results_dir,
                    "recurrent-primary-tumor-enriched-oncofetal-splice-junctions-annotated.tsv.gz"))
```

Print session info

```{r}
sessionInfo()
```
